plyranges: a grammar for transforming genomics data

### 2018-05-04 paper.Rmd
some tweaks to the wording in the algebra operation section

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. The parallel operations map to infix operators, which we
surround with _%_ symbols in accordance with R syntax rules.  For
example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In constrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing whereever there was overlap between a range in _x_
and a range in _y_. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within the arithmetic, restriction and aggregation
expressions.

Our algebra recasts finding overlaps or nearest neighbours between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects, a query and a subject. The join operator
is relational in the sense that metadata from the query and subject
ranges is retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figures 
\ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

>>>>

As with other genomic grammars, we define set operations that treat
ranges as sets of integers, including _intersect_, _union_,
_difference_, and _complement_. There are two sets of these: parallel
and merging. The parallel operations map to infix operators, which we
surround with _%_ symbols, a convention borrowed from R syntax.  For
example, the parallel intersection (_x %intersect% y_) finds the
intersecting range between _xi_ and _yi_ for _i_ in _1...n_, where _n_
is the length of both _x_ and _y_. In constrast, the merging
intersection (_intersect\_ranges(x, y)_) returns a new set of disjoint
ranges representing whereever there was overlap between a range in _x_
and a range in _y_. We use the infix syntax for the parallel
operations, since it is the conventional syntax for parallel, binary
operations in R. Finding the parallel union will fail when two ranges
have a gap, so we introduce a _span_ operator that takes the union
while filling any gap. The _complement_ operation is unique in that it
is unary. It finds the regions not covered by any of the ranges in a
single set. Closely related is the _between_ parallel operation, which
finds the gap separating _xi_ and _yi_. The binary operations are
callable from within arithmetic, restriction and aggregation
expressions.

To support merging, our algebra recasts finding overlaps or 
nearest neighbours between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects, a query and a subject. The join operator
is relational in the sense that metadata from the query and subject
ranges is retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figures 
\ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).
